Bounds, Semantics/Pragmatics-Divide, and Evolution

### 2016-07-05 paper/paper.tex
fragments for paper

\documentclass[11pt]{article}

\usepackage[backgroundcolor=white,textsize=small,linecolor=red]{todonotes}

\usepackage[margin=30mm]{geometry}
\usepackage{natbib}

%--------------------
% Math
\usepackage[fleqn]{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{multicol}
\usepackage{graphicx}

\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}}

\newcommand{\hloranj}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{min}}\;}
\newcommand{\sbna}{\exists\lnot\forall}

%--------------------
%\usepackage{gb4e}
%
%\usepackage{setspace}
%\onehalfspacing
%
%\usepackage{soul} %underlining
%\setuldepth{a}

%\usepackage{inconsolata}

%-------------------

@@

\author{%\bf NAME1 and NAME2\\
    \today
}

@@

\begin{document}
\maketitle

\section{Short exposition}
The plan is to expand our work on the semantics/pragmatics divide by means of a more involved \& principled study on the lack of upper-bounds in the literal meaning of scalar expressions. The main idea of this extension, other than providing a more detailed treatment as well as a comparison with iterated learning, is to do away with the assumption of communal learning. Potential extensions include more scalar pairs, the addition of an inductive bias towards uniform lexica across scalar pairs, and a variation of state frequencies.

\section{Differences to CogSci setup.}
\subparagraph{(I) No communal learning.} Only parental learning is considered

\subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state. A datum $d$ was a sequence of length $k$ of members of $O$. Learners witnessed such data sequences. Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$. The main difference is that now some $d$ do not provide any production information for some states.

@@

\subparagraph{(IV) Parametrized learning} $Q_{ij} \propto [\sum_d P(d|t_i) P(t_j|d)]^l$, where $l =1$ corresponds to probability matching and, as $l$ increases towards infinity, to MAP.

\section{Further modifications we may want to keep in mind (but currently don't make use of)}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

@@

\section{General setup}
\subsection{Part 1: Signaling.}

We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct player types $t \in T$. $|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix. These are listed in Table \ref{tab:lexica}.

\vspace{-0.15cm}
\begin{flalign}
&R_{0}(s|m;L) \propto pr(s) L_{sm}\label{litl}\\
&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{lits}\\
&R_{1}(s|m;L) \propto pr(s) S_{0}(m|s;L) \label{pragl}\\
&S_{1}(m|s;L) \propto  \exp(\lambda \; R_{0}(s|m;L)^\alpha) \label{prags}
\end{flalign}

\subsection{Part 2: Mutation \& replication.}
The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$. $p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type. 
\begin{itemize}
    \item $f_i = \sum_j x_j U(x_i,x_j)$
    \item $\Phi = \sum_i x_i f_i$
    \item $Q_{ij} \propto [\sum_d P(d|t_i) \; P(t_j|d)]^l$, where $P(t_j|d) \propto p^\star(t_j) P(d|t_j)$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ... \tuple{s_k, m_l}}, and $l \geq 1$ is a learning parameter.
	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
\end{itemize}

\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.

>>>>

\documentclass[a4paper]{article}

\usepackage{geometry}
\usepackage{natbib}

%--------------------
\usepackage{gb4e}
\noautomath

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{stmaryrd}
%\usepackage{multicol}
%\usepackage{graphicx}
\usepackage{color}
%\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}}

\newcommand{\hl}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange
%\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}
%\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{min}}\;}
%\newcommand{\sbna}{\exists\lnot\forall}

%--------------------
%
%\usepackage{setspace}
%\onehalfspacing
%
%-------------------

@@

\author{%\bf NAME1 and NAME2\\
    ( -- draft \today --- )
}

@@

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
While there is disagreement on where its boundaries lie, a distinction between semantics and pragmatics is commonly agreed upon by linguists. Broadly put, semantics can be viewed as concerning the literal, truth-conditional, meaning of expressions whereas pragmatics concerns information conveyed by expressions that goes beyond its literal meaning. Consider, for instance, the following sentence

\begin{exe}
\ex SI sentence \begin{xlist}
	\ex Numeral n-1
	\ex SI
\end{xlist}
\end{exe}

%Note first, that the literal meaning of {\em NUMERAL} is compatible with that of more informative or ``stronger'' alternatives such as {\em n+1, n+1} and so on.

\paragraph{Semantics-pragmaits distinction and where the puzzle arises.}

\paragraph{Description of plan to lay out model and test it on upper-bounds.}

\section{Cultural transmission \& linguistic pressures}

\paragraph{Overview of generally assumed linguistic pressures.} \hl{Focus has been put mainly on one over the other, even though it is mutually accepted that both are important. We involve both and connect this research to the rational probabilistic interlocutors literature}

\subsection{Probabilistic models of language use}

\paragraph{Language (fragments).}

\begin{flalign}
&R_{0}(s|m;L) \propto pr(s) L_{sm}\label{litl}\\
&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{lits}\\
&R_{1}(s|m;L) \propto pr(s) S_{0}(m|s;L) \label{pragl}\\
&S_{1}(m|s;L) \propto  \exp(\lambda \; R_{0}(s|m;L)^\alpha) \label{prags}
\end{flalign}

\subsection{Expressiveness}

\subsection{Learnability}
\subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state. A datum $d$ was a sequence of length $k$ of members of $O$. Learners witnessed such data sequences. Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$. The main difference is that now some $d$ do not provide any production information for some states.

@@

\subparagraph{(IV) Parametrized learning} $Q_{ij} \propto \sum_d P(d|t_i) F(t_j,d)$, where $F(t_j,d) \propto P(t_j|d)^l$ and $l =1$ corresponds to probability matching and, as $l$ increases towards infinity, to MAP.

The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$. $p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type. 
\begin{itemize}
    \item $f_i = \sum_j x_j U(x_i,x_j)$
    \item $\Phi = \sum_i x_i f_i$
    \item $Q_{ij} \propto [\sum_d P(d|t_i) \; P(t_j|d)]^l$, where $P(t_j|d) \propto p^\star(t_j) P(d|t_j)$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ... \tuple{s_k, m_l}}, and $l \geq 1$ is a learning parameter.
	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
\end{itemize}

\subsection{Summary}
\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.

\section{Upper-bounds in weak scalar alternatives}
\hl{phenomenon description}

\subsection{Analysis}

\paragraph{Languages.} We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct player types $t \in T$. $|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix. These are listed in Table \ref{tab:lexica}.

\section{Discussion}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

@@

\section{Conclusion}
